---
title: "LA_EDA"
output: html_document
date: "2024-10-26"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
 
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

**SMART QUESTIONS**

1) Which neighborhoods in Los Angeles have experienced the most significant increases in crime rates from 2020 to the present, and are these trends consistent for the same types of crimes year over year?
2) How have theft, robbery, and homicide rates in Los Angeles changed from 2020 to the present, and which of these categories shows the highest overall density of crime during this time?
3) How does the type or frequency of crimes in Los Angeles vary by victim descent, and are there significant geographic patterns (based on latitude and longitude) associated with specific victim groups?
4) Which type of crime in Los Angeles is most frequently associated with weapon use, and what specific types of weapons are most commonly involved in those incidents over the last 2 years?
5) How have the top three most common crimes from 2020 to the present been distributed across the top five areas where they are most frequently committed in Los Angeles, and are these trends increasing or decreasing in each area from 2020 to 2024?


```{r}
crime = read.csv("Crime_Data_from_2020_to_Present.csv")
head(crime)
```

```{r}
nrow(crime)
ncol(crime)
```

The dataset contains 28 columns and 986500 rows.
1) DR_NO - Division of Records Number: Official file number.
2) Date Rptd - MM/DD/YYYY
3) DATE OCC - MM/DD/YYYY
4) TIME OCC - In 24 hour military time
5) AREA - The LAPD has 21 Community Police Stations referred to as Geographic Areas within the department. These Geographic Areas are sequentially numbered from 1-21.
6) AREA NAME - The 21 Geographic Areas or Patrol Divisions are also given a name designation that references a landmark or the surrounding community that it is responsible for.
7) Crm Cd - Indicates the crime committed. Crime Code 1 is the primary and most serious one. Crime Code 2, 3, and 4 are respectively less serious offenses. Lower crime class numbers are more serious.
8) Crm Desc - Indicates the crime description
9) Vict Age - Age of victim
10) Vict Sex - 	F : Female M : Male X : Unknown
11) Vict Descent - 	Descent Code: A - Other Asian B - Black C - Chinese D - Cambodian F - Filipino G - Guamanian H - Hispanic/Latin/Mexican I - American Indian/Alaskan Native J - Japanese K - Korean L - Laotian O - Other P - Pacific Islander S - Samoan U - Hawaiian V - Vietnamese W - White X - Unknown Z - Asian Indian
12) Weapon Desc - Defines the Weapon Used Code provided.
13) Location - Street address of crime incident rounded to the nearest hundred block to maintain anonymity.
14) LAT - Latitude
15) LON - Longtitude


```{r}
na_count <- colSums(is.na(crime))
print("\nNA Count per Column:")
print(na_count)


```

```{r}
new_crime <- crime[, colSums(is.na(crime)) == 0]
head(new_crime)

nrow(new_crime)
ncol(new_crime)
```
```{r}
# Feature Selection based on Smart Questions

cols_to_remove <- c(
   "Mocodes", "Rpt.Dist.No", "Part.1.2", 
  "Premis_cd","Premis.Desc", "Status", "Status.Desc","Cross.Street"
)

# Drop the specified columns
crime_data <- new_crime[, !(names(new_crime) %in% cols_to_remove)]


print("Data after removing unnecessary columns:")
print(names(crime_data))
```

```{r}

library(dplyr)
library(ggplot2)
library(lubridate)

# Ensure date is in correct format
crime_data$DATE.OCC <- as.Date(crime_data$DATE.OCC, format = "%m/%d/%Y")


# Identify top 3 crimes overall from 2020 to present
top_crimes <- crime_data %>%
  filter(year(DATE.OCC) >= 2020) %>%
  group_by(Crm.Cd, Crm.Cd.Desc) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(3)
crime_summary <- crime_data %>%
  filter(year(DATE.OCC) >= 2020) %>%
  group_by(AREA.NAME) %>%
  summarise(
    total_crime = n(),
    top3_crime = sum(Crm.Cd %in% top_crimes$Crm.Cd)
  ) %>%
  mutate(top3_pct = top3_crime / total_crime * 100)
top_areas <- crime_summary %>%
  arrange(desc(top3_crime)) %>%
  head(5)
# Identify top 5 areas with highest counts of top 3 crimes


# Filter data for top 3 crimes from 2020 onwards
crime_filtered <- crime_data %>%
  filter(year(DATE.OCC) >= 2020, Crm.Cd %in% top_crimes$Crm.Cd)
```



Question 5:
How have the top three most common crimes from 2020 to the present been distributed across the top five areas where they are most frequently committed in Los Angeles, and are these trends increasing or decreasing in each area from 2020 to 2024?

```{r}

library(dplyr)
library(ggplot2)
library(lubridate)

# Ensure date is in correct format
crime_data$DATE.OCC <- as.Date(crime_data$DATE.OCC, format = "%m/%d/%Y")

# Calculate crime counts by year for top 3 crimes in top 5 areas
crime_counts <- crime_data %>%
  filter(year(DATE.OCC) >= 2020, 
         AREA.NAME %in% top_areas$AREA.NAME, 
         Crm.Cd %in% top_crimes$Crm.Cd) %>%
  group_by(AREA.NAME, Year = year(DATE.OCC), Crm.Cd.Desc) %>%
  summarise(crime_count = n()) %>%
  ungroup()

# Calculate percent change from 2020 benchmark
crime_change <- crime_counts %>%
  group_by(AREA.NAME, Crm.Cd.Desc) %>%
  mutate(
    benchmark_2020 = crime_count[Year == 2020],
    percent_change = (crime_count - benchmark_2020) / benchmark_2020 * 100
  ) %>%
  ungroup()

# Create a line graph showing percent change from 2020 benchmark for each area
p <- ggplot(crime_change, aes(x = Year, y = percent_change, color = Crm.Cd.Desc, group = Crm.Cd.Desc)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  facet_wrap(~ AREA.NAME, ncol = 2) +
  theme_minimal() +
  labs(title = "Percent Change in Top 3 Crimes from 2020 Benchmark - Top 5 Areas",
       x = "Year",
       y = "Percent Change from 2020",
       color = "Crime Type") +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, size = 10)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

print(p)

# Print crime counts and percent changes for each area
print(crime_change)
```



Question 1

Which neighborhoods in Los Angeles have experienced the most significant increases in crime rates from 2020 to 2023 with particular attention to demographic factors (race and sex of victims), and what insights can be drawn from Area3's significant increase in crimes?

```{r}

library(dplyr)
library(tidyr)
library(knitr)
library(lubridate)

# Assuming crime_data is your original dataset and DATE.OCC is a date column
crime_data_filtered <- crime_data %>%
  filter(year(DATE.OCC) < 2024)

# Create a summary table with year-over-year percentage changes
crime_summary <- crime_data_filtered %>%
  mutate(Year = year(DATE.OCC)) %>%
  filter(Year >= 2020) %>%
  group_by(AREA.NAME, Year) %>%
  summarise(crime_count = n(), .groups = "drop") %>%
  group_by(AREA.NAME) %>%
  mutate(
    previous_year_count = lag(crime_count),
    yoy_percent_change = (crime_count - previous_year_count) / previous_year_count * 100
  ) %>%
  ungroup()

# Create a wide format table
crime_table <- crime_summary %>%
  select(AREA.NAME, Year, yoy_percent_change) %>%
  pivot_wider(
    names_from = Year,
    values_from = yoy_percent_change,
    names_prefix = "percent_change_"
  )

# Print the table
kable(crime_table, format = "markdown", digits = 2)

```

```{r}

library(dplyr)
library(lubridate)

# Assuming crime_data is your original dataset and DATE.OCC is a date column
crime_summary <- crime_data %>%
  mutate(Year = year(DATE.OCC)) %>%
  filter(Year %in% c(2020, 2023)) %>%
  group_by(AREA.NAME, Year) %>%
  summarise(crime_count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Year, values_from = crime_count) %>%
  mutate(percent_change = (`2023` - `2020`) / `2020` * 100) %>%
  arrange(desc(percent_change))

# Print the summary
print(crime_summary)



```


```{r}
library(dplyr)
library(lubridate)

# Calculate crime counts and percent change by area and crime type
crime_analysis <- crime_data %>%
  filter(year(DATE.OCC) %in% c(2020, 2023), 
         AREA.NAME %in% top_areas$AREA.NAME) %>%
  group_by(AREA.NAME, Crm.Cd.Desc, Year = year(DATE.OCC)) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Year, 
              values_from = count, 
              names_prefix = "count_") %>%
  mutate(
    percent_change = ((count_2023 - count_2020) / count_2020) * 100,
    absolute_change = count_2023 - count_2020
  ) %>%
  filter(!is.na(percent_change)) %>%
  arrange(AREA.NAME, desc(percent_change))

# Get top 3 most increased crimes for each area
top_crimes_by_area <- crime_analysis %>%
  group_by(AREA.NAME) %>%
  filter(percent_change > 0) %>%  # Only include increases
  top_n(3, percent_change) %>%
  arrange(AREA.NAME, desc(percent_change))

# Print results
print(top_crimes_by_area)

```


```{r}
library(dplyr)
library(ggplot2)
library(lubridate)

# Calculate crime changes by Area, Crime Type, Sex, and Race
crime_analysis <- crime_data %>%
  filter(year(DATE.OCC) %in% c(2020, 2023)) %>%
  group_by(AREA.NAME, Crm.Cd.Desc, Vict.Sex, Vict.Descent, Year = year(DATE.OCC)) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Year, 
              values_from = count, 
              names_prefix = "count_") %>%
  mutate(
    percent_change = ((count_2023 - count_2020) / count_2020) * 100,
    absolute_change = count_2023 - count_2020
  ) %>%
  filter(!is.na(percent_change))

# Create multiple visualizations

# 1. Crime Change by Area and Sex
p1 <- ggplot(crime_analysis, 
       aes(x = AREA.NAME, y = percent_change, fill = Vict.Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Crime Change by Area and Victim Sex (2020-2023)",
       x = "Area",
       y = "Percent Change",
       fill = "Victim Sex")

# 2. Crime Change by Area and Race/Descent
p2 <- ggplot(crime_analysis, 
       aes(x = AREA.NAME, y = percent_change, fill = Vict.Descent)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Crime Change by Area and Victim Descent (2020-2023)",
       x = "Area",
       y = "Percent Change",
       fill = "Victim Descent")




# Print all plots
print(p1)
print(p2)

# Create summary table
summary_table <- crime_analysis %>%
  group_by(AREA.NAME, Vict.Sex, Vict.Descent) %>%
  summarise(
    avg_percent_change = mean(percent_change, na.rm = TRUE),
    total_crimes_2020 = sum(count_2020, na.rm = TRUE),
    total_crimes_2023 = sum(count_2023, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_percent_change))

print(summary_table)


```
